A Fully Abstract Game Semantics for Countable Nondeterminism

### 2018-01-31 cones/paper/cones.tex
Finished the paper.

An \emph{arena} is given by a triple $A=(M_A,\lambda_A,\ts_A)$, where
\begin{itemize}
  \item $M_A$ is a countable set of moves,
  \item $\lambda_A\from m_A\to\OP\times\QA$ designates each move as either an \emph{$O$-move} or a \emph{$P$-move}, and as either a \emph{question} or an \emph{answer}.  
    We define $\lambda_A^{OP}=\pr_1\circ\lambda_A$ and $\lambda_A^{QA}=\pr_2\circ\lambda_A$.  
    We also define $\neg\from\OP\times\QA\to\OP\times\QA$ to be the function that reverses the values of $O$ and $P$ while leaving $\QA$ unchanged.
  \item $\ts_A$ is an \emph{enabling relation} between $M_A+\{*\}$ and $M_A$ satisfying the following rules:
    \begin{itemize}
      \item If $a\ts_A b$ and $a\ne b$, then $\lambda_A^{OP}(a)\neq\lambda_A^{OP}(b)$.  
      \item If $*\ts_A a$, then $\lambda_A(a)=OQ$ and $b\nts_A a$ for all $b\in M_A$.
      \item If $a\ts_A b$ and $b$ is an answer, then $a$ is a question.
    \end{itemize}
    We say that a move $a\in M_A$ is \emph{initial} in $A$ if $*\ts_A a$.
\end{itemize}

@@

Given an arena $A$, a \emph{justified string} in $A$ is a sequence $s$ of moves in $A$, together with \emph{justification pointers} that go from move to move in the sequence in such a way that every non-initial move $m$ in $s$ has exactly one justification pointer going back to an earlier move $n$ in $s$ such that $n\ts_A m$.  
We say that $n$ \emph{justifies} $m$.  
It is easy to see that every justified string must begin with an initial move, an hence with an $O$-question.

A \emph{legal play} $s$ is a justified string in $A$ that strictly alternates between $O$-moves and $P$-moves and is such that the corresponding $QA$-sequence formed by applying $\lambda_A^{QA}$ to terms is well-bracketed.
We write $L_A$ for the set of legal plays in $A$.

@@

Our base games will be the games $\bN$ and $\bC$ on the arenas of the same names, where $P_{\bN}=\{\emptyplay,q\}\cup\{qn\suchthat n\in\bN\}$ and $\bC = \{\emptyplay,q,qa\}$.

@@

We follow \cite{mcCHFiniteND} by modelling a strategy as a pair $(T_\sigma,D_\sigma)$, where $T_\sigma$ is a nondeterministic strategy in the usual sense and $D_\sigma$ is a set of $O$-plays after which there is a possibility of divergence.

Tracking divergences explicitly in this way requires some care when we compose strategies.  
Specifically, we need to be able to add new divergences into strategies when they arise through `infinite chattering' or \emph{livelock}.  
For example, the denotation of the term
\[
  M = \Y_{\nat\to\nat}(\lambda f.\lambda n.n;(f n))
  \]
is given by a total strategy, without divergences: namely the strategy $\mu$ with plays of the form
\[
  \begin{array}{ccccccc}
    \mathbf{\bN_1} &   & q & n_1 & q & n_2 & \cdots \\
    \mathbf{\bN_2} & q &   &     &   &     &
  \end{array}\,.
  \]
However, when we compose this strategy with any total strategy for $\bN$ on the left, we expect the resulting strategy to contain divergences, since the term $M \n$ diverges for any $\n$.
Semantically, this corresponds to the fact that we have a legal interaction $q\;q\;n\;q\;n\;\cdots$ with an infinite tail in $\bN_1$; when we perform `hiding' by restricting the interaction to $\bN$, we have no reply to the initial move $q$.

The approach adopted in \cite{mcCHFiniteND} is to check specifically for infinite chattering between strategies $\sigma\from A\implies B$ and $\tau\from B\implies C$ by checking whether the set $\sigma\|\tau$ contains any infinite increasing sequence of plays ending with moves in $B$.  
If there is such a sequence, then it restricts to some $O$-position in $\sigma;\tau$ and we add in a divergence at that position.

This works very satisfactorily for finite nondeterminism, but not at all for countable nondeterminism.  
To see why, consider the term
\[
  N = \Y_{\nat\to\nat\to\nat}(\lambda g.\lambda m n.\IfO m\;\0\;n;(g\;(\pred m)\;n))\wn
  \]
This term first chooses a natural number $m$, and then reads from its input $n$ for a total of $m$ times before eventually returning $\0$.  
Thus, its denotation is the strategy $\nu$ with maximal plays of arbitrary length of the form:
\[
  \begin{array}{cccccccc}
    \bN_1 &   & q & n_1 & \cdots & q & n_m & \\
    \bN_2 & q &   &     &        &   &     & 0
  \end{array}\,.
  \]
Note that this strategy strictly contains the one we considered before, and therefore that the denotation of
\[
  \IfO \wn M N
  \]
has the same denotation as $N$, even though for any $n$, $M\n\not{\mustconverge}$, while $N\n\mustconverge$.
Moreover, if we try to compose $\deno{N}$ with the strategy on $\bN$ that always returns $1$, then we end up with an infinite increasing sequence of positions, which triggers the introduction of a divergence into the composite -- even though no divergence occurs in the evaluation of $N$.

@@

Somehow, the crucial point is that we need to distinguish between terms like $M$, which contain infinite sequences of moves, and terms like $N$, which contain arbitrarily large finite sequences of moves.  
The way that we do this is by making the infinite sequences of moves explicit in our strategies, in the style of \cite{RoscoeCspInfinite} and \cite{LevyGsInfinite}.  
Then the denotation of $M$ will contain an infinite sequence, while the denotation of $N$ will contain arbitrarily long finite sequences, but no infinite sequences.

@@

Note that there is no need to consider separately, as Harmer and McCusker do, divergences that arise through `infinite chattering': in our model, a case of infinite chattering between strategies $\sigma$ and $\tau$ is itself a legal interaction between the two strategies, which is necessarily divergent (because it is infinite) and therefore gives rise to some divergence in $\sigma;\tau$.

@@

Any finite-nondeterminism strategy in the sense of \cite{mcCHFiniteND} may be interpreted as a complete strategy by enlarging it with all its infinite limiting plays.  
However, when we introduce countable nondeterminism, we introduce the possibility of strategies that are not complete.  
For example, the strategy $\nu$ that we mentioned above has an infinite increasing sequence of plays $q0\prefix q0q0\prefix \cdots$, but has no infinite play corresponding to its limit.  
Nonetheless, we do not want to allow arbitrary strategies: for example, the strategy $\mu$ above should include the infinite play $qq0q0\dots$; the strategy $\mu^\circ$ formed by removing this infinite play has no meaning in our language.  
Indeed, if we compose $\mu^\circ$ with the strategy $\0$ for $\bN$ on the left, then the resulting strategy does not satisfy diverge-or-reply.
The difference with $\nu$ is that every play $qq0\cdots q0\in T_\nu$ may be completed in $\nu$ by playing the move $0$ on the right.
In other words, $\nu$ is the union of complete strategies, while $\mu^\circ$ is not.

@@

We need to show that the composition of locally complete strategies is locally complete.  
Note that this does not hold for \emph{complete} strategies: for example, our term $N$ above can be written as $N'\;\wn$, where $N'$ is a deterministic term with complete denotation $\nu'$.  
Then we have $\nu=\top_{\bN};\nu'$, but $\nu$ is not complete.
However, we can show that the composition of \emph{deterministic} complete strategies is complete; since a locally complete strategy may always be written as the union of complete deterministic strategies, this is sufficient to show that the composition of locally complete strategies is locally complete.

@@

In fact, the proof of associativity of composition is pretty much the same in our model as it is in any other model of game semantics.  
However, it is worth saying a few words about it, since the model obtained by naively extending the Harmer-McCusker model to unbounded nondeterminism does not have an associative composition.  
The point is that this is not really a problem with associativity, but rather that this naive model gives the wrong result for the composition of strategies with infinite nondeterminism.
For example, if $\nu$ is the strategy we defined above, and $\0$ is the `constant $0$' strategy on $\bN$, then $\0;\nu$ has a divergence in the naive model, because the strategies $\0$ and $\nu$ appear to be engaged in infinite chattering.  
In our model, we have fixed that problem, because the strategy $\nu$ contains no infinite plays, and so no divergences arise in the composition.

@@

This is very similar to the construction of a co-Kleisli category for a linear exponential comonad, but certain technical issues relating to well-openedness prevent us from displaying it in this way.

@@

Most full abstraction results go via a definability result that says that all \emph{compact} strategies are definable \cite{curienFullAbstraction}.
However, deducing full abstraction from compact definability makes essential use of continuity properties that are absent when we deal with countable nondeterminism.  
We will therefore need to appeal to a stronger result -- that of \emph{universality}, which states that \emph{every} strategy is definable.  
Clearly, universality does not hold for any of our categories of games, since there are many non-computable functions $\bN\to\bN$.  
However, Hyland and Ong proved in \cite{hoPcf} that every \emph{recursively presentable} innocent strategy is PCF-definable.

@@

We shall tacitly assume that the games in our category $\G_{ND}$ come with an enumeration so that we can reason about recursive strategies later.  
For any game that is the denotation of an Idealized Algol type, this enumeration will be the natural one as described above.

@@

Our definability results will hinge on a \emph{factorization theorem}, showing that every nondeterministic strategy may be written as the composition of a deterministic strategy with the nondeterministic `oracle' $\top_{\bN}$.  
We can then deduce definability from definability in the model of deterministic Idealized Algol.

@@

  We first assume that the strategy $\sigma$ is complete.
  Then the strategy $\Det(\sigma)$ is very easy to describe.  
  For each $O$-position $sa\in T_\sigma$, we have some set $B$ of possible replies to $sa$, which we order as $b_1,b_2,\cdots$, where $\code_A(b_1)<\code_A(b_2)<\cdots$.  
  We insert a request to the oracle for a natural number; then, depending on her answer $j$, we play the next move as follows:
  \begin{itemize}
    \item If $0 < j \le \code_A(b_1)$, then play $b_1$.
    \item If $\code_A(b_n) < j \le \code_A(b_{n+1})$ then play $b_{n+1}$.
    \item If $j = 0$ and $sa\in D_\sigma$, then play nothing, and put the resulting play inside $D_{\Det(\sigma)}$.  
      Otherwise, play $b_1$.
  \end{itemize}
  Lastly, we close under limits to make the strategy $\Det(\sigma)$ complete.  
  $\Det(\sigma)$ is clearly nondeterministic.  
  Checking that $\top_{\bN};\Det(\sigma)=\sigma$ is easy for finite plays; for infinite plays, it follows by completeness of $\sigma$.

@@

\begin{proposition}[Computational Adequacy]
  $M\converges\skipp$ if and only if $qa\in T_{\deno M}$.  
  $M\mustconverge$ if and only if $D_{\deno M}=\emptyset$.
  \label{prop:adequacy}
\end{proposition}

Traditional proofs of computational adequacy using logical relations make essential use of the continuity of composition with respect to a natural ordering on strategies (see, for example, \cite{mcCHFiniteND} and \cite{RusssThesis} for the finite nondeterminism case).  
In our case, since composition is not continuous in the language itself, we cannot use this technique.
In order to prove adequacy, we use a new technique that involves using a deterministic stateful construction to model the nondeterminism inside a deterministic world in which continuity holds.  
We shall return to the concept of an \emph{evaluation} of a term as a sequence of natural numbers that we use to replace the constant $\wn$.

@@

\begin{proof}
  Let $n_1,\dots,n_k,d$ be a finite sequence of natural numbers.  
  We define an Idealized Algol term $N_{n_1,\dots,n_k,d}\from (\nat \to \com) \to \com$ to be the following.
  \[
    \lambda f.\new_\nat (\lambda v.f(v\coloneqq (suc\;\deref v); \case_{k+1}\;\deref v\;\Omega\;n_1 \cdots n_k d))\,.
    \]
  Here, $\case_{k+1}\;a\;n_0\;\cdots\;n_k\;d$ is a new shorthand that evaluates to $n_i$ if $a$ evaluates to $i$, and evaluates to $d$ if $a$ evaluates to $j> k$.
  This term adds an extra variable $v$ to the program; each time $f$ is called, it increments the value of $v$ and maps its value on to one of the $n_i$.

@@

  Digging into the construction of $\new$ within Idealized Algol, as given in \cite{SamsonGuyIAPassive}, we see that for any term $F$ of type $\nat\to\com$ the denotation of $N_{n_1,\dots,n_k,d} F$ is given by the composite
  \[
    I \xrightarrow{\cell_{0}}
    \oc \Var \xrightarrow{\deno{\lambda v.v\coloneqq (\suc\;\deref v);\case_{k+1}\;\deref v\;\Omega\;n_1\cdots n_k d}}
    \oc\bN \xrightarrow{\deno F}
    \bC\,.
    \]
  We set $\sigma_\pi$ to be the composite of the left two arrows.  
  Observe that $\sigma_\pi$ is the strategy with unique maximal infinite play as follows.
  \[
    q\;n_1\;\cdots\;q\;n_k\;q\;d\;q\;d\;\cdots
    \]
  Setting $F = \lambda n.C[n]$, we see that $\deno{F}=\sigma_M$.  
  So, by adequacy for the Idealized Algol model, $qa\in T_{\sigma_\pi;\sigma_M}$ if and only if $N_{n_1,\dots,n_k,d} (\lambda n.C[n])\converges\skipp$, which is the case if and only if $M\converges\skipp$ along the evaluation $\pi$.
  Similarly, $D_{\sigma_\pi;\sigma_M}\neq\emptyset$ if and only if $N_{n_1,\dots,n_k,d}(\lambda n.C[n])$ diverges, which is equivalent to saying that $M$ diverges along the evaluation $\pi$.

@@

\begin{proof}
  Since $\sigma$ is total and deterministic, it must have a maximal infinite play $s_\sigma$ of the form
  \[
    q\;m_1\;q\;n_2\;\cdots\,,
    \]
  where $m_1,m_2,\dots$ is some infinite sequence of natural numbers.  
  If the strategy $\sigma_M$ contains some play $\s$ such that $\s\vert_{\oc\bN}=s$, then $\sigma=\sigma_\pi$ for some infinite evaluation $\pi$ of $M$.  
  Otherwise, let $t$ be the maximal sub-play of $s$ such that $\s\vert_{\oc\bN}=t$ for some $\s\in\sigma_M$.  
  Then, if we replace $\sigma$ with the strategy $\sigma'$ that plays according to $t$ and subsequently plays $q\;d\;q\;d\;\cdots$ for our fixed value $d$, then we will have $\sigma';\sigma_M=\sigma;\sigma_M$.  
  Moreover, $\sigma'=\sigma_\pi$ for some evaluation $\pi$ of $M$.

  Now suppose that there exists $\sigma\from\oc\bN$ such that $qa\in T_{\sigma;\sigma_M}$.  
  By the proof of Lemma \ref{lem:soundness-lemma}, we may assume that $\sigma=\sigma_\pi$ for some evaluation $\pi$ of $M$.  
  Therefore, $qa\in T_{\sigma_\pi;\sigma_M}$, which means that $M\converges\skipp$ along the evaluation $\pi$.  
  The corresponding statement for must convergence follows in the same way.
\end{proof}

@@

The point here is that even though $q$ is not explicitly a divergence in the second case, it is nonetheless impossible to prevent the strategy from eventually reaching a divergence.  
Given a strategy $\sigma$ for a game $A$, we say that $s\in\sigma$ is \emph{unreliable} if player $P$ has a strategy, starting from $s$ and staying at all times within $T_\sigma$, that ensures that we eventually reach some play in $D_\sigma$.  
We then say that a strategy $\sigma$ is \emph{divergence-complete} if every unreliable point of $\sigma$ is contained in $D_\sigma$.  
Every strategy $\sigma$ can clearly be extended to a minimal divergence-complete strategy $\text{dc}(\sigma)$; Murawski's explicit characterization of the intrinsic collapse \cite{MurawskiIntrinsic}, which may be applied to our model, essentially says that $\sigma\oes\tau$ if and only if $\text{dc}(\sigma)=\text{dc}(\tau)$.

@@

Instead, they use a novel type of game semantics in which strategies are given not by sets of plays but by justified \emph{trees} whose vertices are moves.  
The difference is subtle but important: from the point of view of countable nondeterminism it allows us to distinguish between terms that take an arbitrarily large finite number of steps to converge and terms that diverge.  
For example, the denotation of our term $M$ above has an infinite branch in it, while the denotation of $N$ has infinitely many finite branches of arbitrary length.

The main advantage of the tree-based model, however, is that it gives rise to a satisfactory synthetic definition of a \emph{nondeterministic innocent} strategy based on sheaves.  
In fact, Tsukada and Ong do not prove full abstraction for PCF with countable nondeterminism (even though that language may be interpreted within their category), but for PCF with finite nondeterminism.

@@

Instead, we need to look at the combinatorial structure of the strategies.  
We observe that if $\sigma_M\from\oc\bN\to\bC$ is a strategy, then there exists a deterministic strategy $\sigma\from\oc\bN$ such that $qa\in T_{\sigma;\sigma_M}$ if and only if there exists some $\s\in T_{\sigma_M}$ such that $\s\vert_{\bC}=qa$.

@@

\begin{definition}
  Let $\sigma,\tau\from\bN_1\to\bN_2$ be morphisms in $\G_{inn}^\oc$.  
  We say that $\sigma\approx\tau$ if the following two conditions hold.
  \begin{itemize}
    \item For all $n\in\omega$, there exists $\s\in\sigma$ such that $\s\vert_{\bN_2}=qn$ if and only if there exists $\t\in\tau$ such that $\t\vert_{\bN_2}=qn$.  
    \item $\sigma$ is winning if and only if $\tau$ is winning.
  \end{itemize}
\end{definition}

>>>>

An \emph{arena} is given by a triple $A=(M_A,\lambda_A,\ts_A)$, where
\begin{itemize}
  \item $M_A$ is a countable set of moves,
  \item $\lambda_A\from M_A\to\OP\times\QA$ designates each move as either an \emph{$O$-move} or a \emph{$P$-move}, and as either a \emph{question} or an \emph{answer}.  
    We define $\lambda_A^{OP}=\pr_1\circ\lambda_A$ and $\lambda_A^{QA}=\pr_2\circ\lambda_A$.  
    We also define $\neg\from\OP\times\QA\to\OP\times\QA$ to be the function that reverses the values of $O$ and $P$ while leaving $\QA$ unchanged.
  \item $\ts_A$ is an \emph{enabling relation} between $M_A+\{*\}$ and $M_A$ satisfying the following rules:
    \begin{itemize}
      \item If $a\ts_A b$ and $a\ne b$, then $\lambda_A^{OP}(a)\neq\lambda_A^{OP}(b)$.  
      \item If $*\ts_A a$, then $\lambda_A(a)=OQ$ and $b\nts_A a$ for all $b\in M_A$.
      \item If $a\ts_A b$ and $b$ is an answer, then $a$ is a question.
    \end{itemize}
    We say that a move $a\in M_A$ is \emph{initial} in $A$ if $*\ts_A a$.
\end{itemize}

@@

Given an arena $A$, a \emph{justified string} in $A$ is a sequence $s$ of moves in $A$, together with \emph{justification pointers} that go from move to move in the sequence.  
The justification pointers must be set up in such a way that every non-initial move $m$ in $s$ has exactly one justification pointer going back to an earlier move $n$ in $s$ such that $n\ts_A m$.  
We say that $n$ \emph{justifies} $m$.  
It is easy to see that every justified string must begin with an initial move, an hence with an $O$-question.

A \emph{legal play} $s$ is a justified string in $A$ that strictly alternates between $O$-moves and $P$-moves and is such that the corresponding $QA$-sequence formed by applying $\lambda_A^{QA}$ to moves is well-bracketed.
We write $L_A$ for the set of legal plays in $A$.

@@

Our base games will be the games $\bN$ and $\bC$ on the arenas of the same names, where $P_{\bN}=\{\emptyplay,q\}\cup\{qn\suchthat n\in\bN\}$ and $P_{\bC} = \{\emptyplay,q,qa\}$.

@@

To fix this problem, we follow \cite{mcCHFiniteND} by modelling a strategy as a pair $(T_\sigma,D_\sigma)$, where $T_\sigma$ is a nondeterministic strategy in the usual sense and $D_\sigma$ is a set of $O$-plays after which there is a possibility of divergence.

Tracking divergences explicitly in this way requires some care when we compose strategies using `parallel composition plus hiding'.
Specifically, we need to be able to add new divergences into strategies when they arise through `infinite chattering' or \emph{livelock}.  
For example, the denotation of the term
\[
  M = \Y_{\nat\to\nat}(\lambda f.\lambda n.n;(f n))
  \]
is given by a total strategy, without divergences: namely the strategy $\mu$ with plays of the form shown in Figure \ref{fig:mn-plays}(a).
\begin{figure}
  \begin{mathpar}
    a) \and
    \begin{array}{cc}
      \mathbf{\bN_1} & \mathbf{\bN_2} \\
                     & q \\
                  q  &   \\
                 n_1 &   \\
                  q  &   \\
                 n_2 &   \\
              \vdots &   \\
              {}     & {}
    \end{array}
    \and b) \and
    \begin{array}{cc}
      \mathbf{\bN_1} & \mathbf{\bN_2} \\
                     & q \\
                  q  &   \\
                 n_1 &   \\
              \vdots &   \\
                  q  &   \\
                 n_m &   \\
                     & 0 \\
    \end{array}
  \end{mathpar}
  \caption{Finite plays alone are not sufficient to distinguish between terms of a language with countable nondeterminism.}
  \label{fig:mn-plays}
\end{figure}
However, when we compose this strategy with any total strategy for $\bN$ on the left, we expect the resulting strategy to contain divergences, since the term $M \n$ diverges for any $\n$.
Semantically, this corresponds to the fact that we have a legal interaction $q\;q\;n\;q\;n\;\cdots$ with an infinite tail in $\bN_1$; when we perform `hiding' by restricting the interaction to $\bN$, we have no reply to the initial move $q$.

The approach adopted in \cite{mcCHFiniteND} is to check specifically for infinite chattering between strategies $\sigma\from A\implies B$ and $\tau\from B\implies C$ by checking whether there is an infinite increasing sequence of interactions between $\sigma$ and $\tau$ with an infinite tail in $B$.  
If there is such a sequence, then it restricts to some $O$-position in $\sigma;\tau$ and we add in a divergence at that position.

This works very satisfactorily for finite nondeterminism, but not at all for countable nondeterminism.  
To see why, consider the term
\[
  N = \Y_{\nat\to\nat\to\nat}(\lambda g.\lambda m n.\IfO m\;\0\;n;(g\;(\pred m)\;n))\wn
  \]
This term first chooses a natural number $m$, and then reads from its input $n$ for a total of $m$ times before eventually returning $\0$.  
Thus, its denotation is the strategy $\nu$ with maximal plays of arbitrary length of the form shown in Figure \ref{fig:mn-plays}(b).
Note that this strategy strictly contains the strategy $\mu$ that we considered before, and therefore that the denotation of
\[
  \IfO \wn M N
  \]
has the same denotation as $N$, even though for any $n$, $M\n\not{\mustconverge}$, while $N\n\mustconverge$.
Moreover, if we try to compose $\deno{N}$ with the strategy on $\bN$ that always returns $1$, then we end up with an infinite increasing sequence of positions, which triggers the introduction of a divergent play into the composite strategy -- even though no divergence occurs in the evaluation of $N$.

@@

Somehow, the crucial point is that we need to distinguish between terms like $M$, which contain infinite sequences of moves, and terms like $N$, which contain arbitrarily large finite sequences of moves.  
The way that we do this is by making the infinite sequences of moves explicit in our strategies, in the style of \cite{RoscoeCspInfinite} and \cite{LevyGsInfinite}.  
When we use this technique, the denotation of $M$ will contain an infinite sequence, while the denotation of $N$ will contain arbitrarily long finite sequences, but no infinite sequences.

@@

Note that there is no need to consider separately, as Harmer and McCusker do, divergences that arise through `infinite chattering': in our model, a case of infinite chattering between strategies $\sigma$ and $\tau$ is itself an (infinite) legal interaction between the two strategies, which is necessarily divergent (because it is infinite) and therefore gives rise to some divergence in $\sigma;\tau$.

@@

Any finite-nondeterminism strategy in the sense of \cite{mcCHFiniteND} may be interpreted as a complete strategy by enlarging it with all its infinite limiting plays.  
However, when we introduce countable nondeterminism, we also introduce strategies that are not complete.  
For example, the strategy $\nu$ that we mentioned above has an infinite increasing sequence of plays $q0\prefix q0q0\prefix \cdots$, but has no infinite play corresponding to its limit.  
Nonetheless, we do not want to allow arbitrary strategies: for example, the strategy $\mu$ above should include the infinite play $qq0q0\dots$; the strategy $\mu^\circ$ formed by removing this infinite play has no meaning in our language.  
Indeed, if we compose $\mu^\circ$ with the strategy $\0$ for $\bN$ on the left, then the resulting strategy does not satisfy diverge-or-reply.
The difference with $\nu$ is that every play $qq0\cdots q0\in T_\nu$ may be completed in $\nu$ by playing the move $0$ on the right.
In other words, $\nu$ is the union of complete strategies, while $\mu^\circ$ is not.

@@

We need to show that the composition of locally complete strategies is locally complete.  
Note that the composition of \emph{complete} strategies is not necessarily complete: for example, our term $N$ above can be written as $N'\;\wn$, where $N'$ is a deterministic term with complete denotation $\nu'$.  
Then we have $\nu=\top_{\bN};\nu'$, but $\nu$ is not complete.
However, we can show that the composition of \emph{deterministic} complete strategies is complete; since a locally complete strategy may always be written as the union of complete deterministic strategies, this is sufficient to show that the composition of locally complete strategies is locally complete.

@@

In fact, the proof of associativity of composition is pretty much the same in our model as it is in any other model of game semantics.  
However, it is worth saying a few words about it, since the model obtained by naively extending the Harmer-McCusker model to unbounded nondeterminism does not have an associative composition.  
The point is that there is not really a problem with associativity itself, but rather that this naive model gives the wrong result for the composition of strategies with infinite nondeterminism.
For example, if $\nu$ is the strategy we defined above, and $\0$ is the `constant $0$' strategy on $\bN$, then $\0;\nu$ has a divergence in the naive model, because the strategies $\0$ and $\nu$ appear to be engaged in infinite chattering.  
In our model, we have fixed that problem, because the strategy $\nu$ contains no infinite plays, and so no divergences arise in the composition.

@@

This is very similar to the construction of a co-Kleisli category for a linear exponential comonad, but certain technical issues relating to well-openedness prevent us from presenting it in this way.

@@

Most full abstraction results go via a definability result that says that all \emph{compact} strategies are definable \cite{curienFullAbstraction}.
However, deducing full abstraction from compact definability makes essential use of continuity properties that are absent when we deal with countable nondeterminism.  
We will therefore need to appeal to a stronger result -- that of \emph{universality}, which states that \emph{every} strategy is definable.  
Clearly, universality does not hold for any of our categories of games -- for example, there are many non-computable functions $\bN\to\bN$.  
However, Hyland and Ong proved in \cite{hoPcf} that every \emph{recursively presentable} innocent strategy is PCF-definable.

@@

We shall tacitly assume that the games in our category $\G_{ND}$ come with an enumeration so that we can reason about recursive strategies later.  
For any game that is the denotation of an Idealized Algol type, this enumeration will be as we have just described.

@@

Our definability results will hinge on a \emph{factorization theorem}, showing that every nondeterministic strategy may be written as the composition of a deterministic strategy with the nondeterministic `oracle' $\top_{\bN}$.  
We can then deduce universality from universality in the model of deterministic Idealized Algol.

@@

  We first assume that the strategy $\sigma$ is complete.
  Then the strategy $\Det(\sigma)$ is very easy to describe.  
  For each $O$-position $sa\in T_\sigma$, we have some set $B$ of possible replies to $sa$, which we order as $b_1,b_2,\cdots$, where $\code_A(b_1)<\code_A(b_2)<\cdots$.  
  We insert a request to the oracle for a natural number; then, depending on her answer $j$, we play the next move as follows:
  \begin{itemize}
    \item If $0 < j \le \code_A(b_1)$, then play $b_1$.
    \item If $\code_A(b_n) < j \le \code_A(b_{n+1})$ then play $b_{n+1}$.
    \item If $j = 0$ and $sa\in D_\sigma$, then play nothing, and put the resulting play inside $D_{\Det(\sigma)}$.  
      Otherwise, play $b_1$.
  \end{itemize}
  Lastly, we close under limits to make the strategy $\Det(\sigma)$ complete.  
  $\Det(\sigma)$ is clearly deterministic.  
  Checking that $\top_{\bN};\Det(\sigma)=\sigma$ is easy for finite plays; for infinite plays, it follows by completeness of $\sigma$.

@@

\begin{proposition}[Computational Adequacy]
  Let $M\from\com$ be a closed term of Idealized Algol with countable nondeterminism.
  $M\converges\skipp$ if and only if $qa\in T_{\deno M}$.  
  $M\mustconverge$ if and only if $D_{\deno M}=\emptyset$.
  \label{prop:adequacy}
\end{proposition}

Traditional proofs of computational adequacy using logical relations make essential use of the continuity of composition with respect to a natural ordering on strategies (see, for example, \cite{mcCHFiniteND} and \cite{RusssThesis} for the finite nondeterminism case).  
In our case, since composition is not continuous in the language itself, we cannot use this technique.
In order to prove adequacy, we use a new technique that involves using a deterministic stateful construction to model the nondeterminism inside a deterministic world in which continuity holds.  
To do this, we shall return to the concept of an \emph{evaluation} $\pi$ of a term as a sequence of natural numbers that we use to replace the constant $\wn$.

@@

\begin{proof}
  Let $n_1,\dots,n_k,d$ be a finite sequence of natural numbers.  
  We define an Idealized Algol term $N_{n_1,\dots,n_k,d}\from (\nat \to \com) \to \com$ to be the following.
  \[
    \lambda f.\new_\nat (\lambda v.f(v\coloneqq (suc\;\deref v); \case_{k+1}\;\deref v\;\Omega\;n_1 \cdots n_k d))\,.
    \]
  Here, $\case_{k+1}\;a\;n_0\;\cdots\;n_k\;d$ is a new shorthand that evaluates to $n_i$ if $a$ evaluates to $i$, and evaluates to $d$ if $a$ evaluates to $j> k$.
  This term adds an extra variable $v$ to the program; each time $f$ is called, it increments the value of $v$ and maps its value to one of the $n_i$.  
  The result is that when $f$ calls its argument, it will receive $n_1$ the first time, $n_2$ the second and so on.

@@

  Digging into the construction of $\new$ within Idealized Algol, as given in \cite{SamsonGuyIAPassive}, we see that for any term $F$ of type $\nat\to\com$ the denotation of $N_{n_1,\dots,n_k,d} F$ is given by the composite
  \[
    I \xrightarrow{\cell_{0}}
    \oc \Var \xrightarrow{\oc\deno{\lambda v.v\coloneqq (\suc\;\deref v);\case_{k+1}\;\deref v\;\Omega\;n_1\cdots n_k d}}
    \oc\bN \xrightarrow{\deno F}
    \bC\,.
    \]
  We set $\sigma_\pi$ to be the composite of the left two arrows.  
  Observe that $\sigma_\pi$ is the strategy with unique maximal infinite play as follows.
  \[
    q\;n_1\;\cdots\;q\;n_k\;q\;d\;q\;d\;\cdots
    \]
  Setting $F = \lambda n.C[n]$, we see that $\deno{F}=\sigma_M$.  
  So, by adequacy for the Idealized Algol model, $qa\in T_{\sigma_\pi;\sigma_M}$ if and only if we have $N_{n_1,\dots,n_k,d} (\lambda n.C[n])\converges\skipp$, which is the case if and only if $M\converges\skipp$ along the evaluation $\pi$.
  Similarly, $D_{\sigma_\pi;\sigma_M}\neq\emptyset$ if and only if $N_{n_1,\dots,n_k,d}(\lambda n.C[n])$ diverges, which is equivalent to saying that $M$ diverges along the evaluation $\pi$.

@@

\begin{proof}
  Since $\sigma$ is total and deterministic, it must have a maximal infinite play $s_\sigma$ of the form
  \[
    q\;m_1\;q\;m_2\;\cdots\,,
    \]
  where $m_1,m_2,\dots$ is some infinite sequence of natural numbers.  
  If the strategy $\sigma_M$ contains some play $\s$ such that $\s\vert_{\oc\bN}=s$, then $\sigma=\sigma_\pi$ for some infinite evaluation $\pi$ of $M$.  
  Otherwise, let $t$ be the maximal sub-play of $s$ such that $\s\vert_{\oc\bN}=t$ for some $\s\in\sigma_M$.  
  Then, if we replace $\sigma$ with the strategy $\sigma'$ that plays according to $t$ and subsequently plays $q\;d\;q\;d\;\cdots$ for our fixed value $d$, we will have $\sigma';\sigma_M=\sigma;\sigma_M$.  
  In either case, $\sigma'=\sigma_\pi$ for some evaluation $\pi$ of the term $M$.

  Now suppose that there exists $\sigma\from\oc\bN$ such that $qa\in T_{\sigma;\sigma_M}$.  
  We may assume that $\sigma=\sigma_\pi$ for some evaluation $\pi$ of $M$.  
  Therefore, $qa\in T_{\sigma_\pi;\sigma_M}$, which means that $M\converges\skipp$ along the evaluation $\pi$.  
  The corresponding statement for must convergence follows in the same way.
\end{proof}

@@

The point here is that even though $q$ is not explicitly a divergence in the second case, it is nonetheless impossible to prevent the strategy from eventually reaching a divergence.

Given a nondeterministic strategy $\sigma$ for a game $A$, we may treat $\sigma$ as a game in its own right (a sub-game of $A$).  
Moreover, for any $s\in T_\sigma$, we have a particular branch of that game in which play starts at $s$.  
We say that $s$ is \emph{unreliable} if player $P$ has a strategy for the game starting at $s$ that ensures that play eventually ends up in $D_\sigma$.

We then say that a strategy $\sigma$ is \emph{divergence-complete} if every unreliable point of $\sigma$ is contained in $D_\sigma$.  
Every strategy $\sigma$ can clearly be extended to a minimal divergence-complete strategy $\text{dc}(\sigma)$; Murawski's explicit characterization of the intrinsic collapse \cite{MurawskiIntrinsic}, which may be applied to our model, essentially says that $\sigma\oes\tau$ if and only if $\text{dc}(\sigma)=\text{dc}(\tau)$.

@@

Instead, they use a novel type of game semantics in which strategies are given not by sets of plays but by justified \emph{trees} whose vertices are moves.  
The difference is subtle but important: from the point of view of countable nondeterminism it allows us to distinguish between terms that take an arbitrarily large finite number of steps to converge and terms that diverge, without needing to track infinitary information.
For example, the denotation of our term $M$ above has an infinite branch in it, while the denotation of $N$ has infinitely many finite branches of arbitrary length.

The main advantage of the tree-based model, however, is that it gives rise to a satisfactory synthetic definition of a \emph{nondeterministic innocent} strategy, based on sheaves.  
The main contribution of their paper is not to prove full abstraction for PCF with countable nondeterminism (even though that language may be interpreted within their category), but for PCF with finite nondeterminism.

@@

Instead, we need to look at the combinatorial structure of the strategies.  
We observe that if $\sigma_M\from\oc\bN\to\bC$ is a strategy, then there exists a deterministic strategy $\sigma\from\oc\bN$ such that $qa\in T_{\sigma;\sigma_M}$ if and only if there exists some $s\in T_{\sigma_M}$ such that $s\vert_{\bC}=qa$.

@@

\begin{definition}
  Let $\sigma,\tau\from\bN_1\to\bN_2$ be morphisms in $\G_{inn}^\oc$.  
  We say that $\sigma\approx\tau$ if the following two conditions hold.
  \begin{itemize}
    \item For all $n\in\omega$, there exists $s\in\sigma$ such that $s\vert_{\bN_2}=qn$ if and only if there exists $t\in\tau$ such that $t\vert_{\bN_2}=qn$.  
    \item $\sigma$ is winning if and only if $\tau$ is winning.
  \end{itemize}
\end{definition}
